Potential Factors of Decreasing Birth Rates in the US since the Great Recession and Beyond

### 2024-02-11 outputs/paper/paper.qmd
Update paper.qmd
check grammar and tone

# Introduction
1) broader context to motivate
2) some detail about what the paper is about
3) a clear gap that needs to be filled
4) what was done
5) what was found
6) why it is important - replacement level fertility
7) the structure of the paper - 1. trend in US Birth Rates 2. Trends in Birth Rates by Population Subgroup 2.1 Five-year age group 2.2 Race and ethinicity

According to the Centers for Disease Control and Prevention, birth rate is defined as "the number of live births per 1,000 population." It is calculated by dividing the number of live births in a population in a year by the midyear resident population. More specifically, for census years, rates are based on unrounded census counts of the resident population as of April 1. For the noncensus years 1981 - 1989, rates are based on the midyear resident population, rounded to thousands (National Center for Health Statistics 2023). In an economical aspect, birth rates are an important factor that decides the growth of a country. In recent years, low birth rate became a conversation topic both nation-wide and worldwide. Many countries are rising this issue of declining birth rates and posing alternative measures to compensate the population drop. For instance, Germany birth rate (births per 1000 people) in 2024 hits 9.3 which is lower than Canada 10.0 and the US 12.0 (). Germany takes an action to actively recruiting skilled workers from abroad and reforming its citizenship law by allowing dual citizenship

* what was done, why and what was found, 3-4 paragraphs

We replicate the paper by Kearney, Levine, and Pardue (2022) with a focus on the following research questions:

* What is the trend in the US birth rates over the period 1980 to 2020?

* How does the birth rates vary by the age group?

* How does the birth rates vary by the race and ethnicity of mothers'?

>>>>

# Introduction

According to the Centers for Disease Control and Prevention, the birth rate is defined as "the number of live births per 1,000 population." This rate is calculated by dividing the number of live births in a population in a year by the midyear resident population. Specifically, for census years, rates are calculated based on the unrounded census counts of the resident population as of April 1st. From an economic perspective, birth rates are an important factor that determines the growth of a country. In recent years, the low birth rate has become a conversation topic both nationwide and worldwide. Many countries are addressing this issue of declining birth rates and proposing alternative measures to compensate for the population drop. For instance, Germany's birth rate (births per 1000 people) in 2024 is 9.3, which is lower than Canada's 10.0 and the US's 12.0 (Macrotrends 2024). Germany is taking action by actively recruiting skilled workers from abroad (Grunau 2023) and reforming its citizenship law to allow dual citizenship (Knight 2024) as some measures to combat declining birth rates and overall population. Finally, the birth rate also serves as an indicator of replacement level fertility, which in developed countries can be taken as requiring an average of 2.1 children per woman (Craig 1994). Therefore, birth rates are poised to become a key global topic in the mid to long term.

Two economists, Melissa S. Kearney and Phillip B. Levine, along with Luke Pardue, a PhD Candidate in Economics, authored a paper addressing the dramatic decline in birth rates in the United States between 2007 and 2020 and proposed possible explanations. They examined the overall reduction in the birth rate among various population groups of women, including teens, Hispanic women, and college-educated white women. While they attributed the main cause of the decline in the early part of the period to the Great Recession, they were unable to identify any other economic, policy, or social factors that have changed since 2007 to support the decline beyond the Great Recession (Kearney, Levine, and Pardue 2022). The objective of this paper is to replicate the figures and tables from the original study and to explore additional factors contributing to the declining birth rates in the US. In addition to examining the impact of the Great Recession, this paper delves into the perspectives of population subgroups, analyzing data in five-year age groups and by race and ethnicity to gain deeper insights.

This paper utilized a replication package of the original study and employed a cleaned data CSV file to reproduce the related figures and tables. This approach enabled us to draw correlations between the dataset and political, economic, and social policy measures that were not covered in the original paper. Furthermore, this paper investigates various socio-cultural aspects, including sexual education, marriage rates, financial stability among women over the age of 30, and the role of technology in healthcare.

-what was found-

First, the paper showcases the trend in US birth rates and discusses the relevant factors along with the figure. It then demonstrates trends in birth rates by population subgroup, focusing on two key areas: five-year age groups and race and ethnicity. Alongside the graphs and tables, the paper correlates various aspects including education, socio-economic factors, and finance to explain the declining birth rates.

We replicate the study by Kearney, Levine, and Pardue (2022) with a focus on the following research questions:

* What is the trend in US birth rates over the period 1980 to 2020?

* How do birth rates vary by age group?

* How do birth rates vary by the race and ethnicity of mothers?
